Manuscript_Github_Pchabaudi_boilerplate

### 2022-08-03 chapter4_results_chabaudi24h/Thesis_Github_Pchabaudi_qusage.Rmd
Make heatmap of qusage results for stages between transmissions, also do qusage analysis for said comparisons

library('readr')
library('dplyr')
library('ggplot2')
library("RColorBrewer")
library('data.table')
library('tibble')
library('stringr')
library('kableExtra')
library('viridis')
library('reshape2')
library('ComplexHeatmap')
library('circlize') # for circular plots and colour
library("factoextra")
library("MatrixGenerics") # for rowMins and rowMax

@@

cir_info <- read_delim('../standard_files/Pchabaudi_pir_info.txt',
           delim = '\t')
cir_id <- cir_info$Geneid

cir_clade_info <- read_delim('../standard_files/Pchabaudi_pir_clades.txt',
            delim = '\t')

#Add chapl_aapl/subfam numbers for easier heatmap drawing later.
cir_info <- cir_info %>% 
  group_by(chapl_aapl_name) %>% 
  mutate(chapl_aapl_name_num = n_distinct(Geneid)) %>% 
  mutate(
    chapl_aapl_name_num = paste0(
      chapl_aapl_name, 
      ' (n = ',
      chapl_aapl_name_num,
      ')'
    )) %>% 
  group_by(subfam) %>% 
  mutate(subfam_num = n_distinct(Geneid)) %>% 
  mutate(
    subfam_num = paste0(
      subfam, 
      ' (n = ',
      subfam_num,
      ')'
    )
  )
```

@@

chab_tpm_avg.list <- list(
  'RMT' = list(
    tpm = chab_tpm_avg.df[,grepl(colnames(chab_tpm_avg.df), 
                                pattern = 'rMT|Geneid')],
    experiment = 'RMT'
  ),
  'SBP' = list(
    tpm = chab_tpm_avg.df[,grepl(colnames(chab_tpm_avg.df), 
                                pattern = 'SBP|Geneid')],
    experiment = 'SBP'
  )
)

@@

Adding in housekeeping genes for comparison:
HSP70 putative  PCHAS_0721000
MSP1 PCHAS_0831300
AMA1 PCHAS_0931000

```{r highest_expressed_stage_heatmap_fig3b}

housekeeping_genes <- data.frame('HSP70' = 'PCHAS_0721000',
                                 'MSP1' = 'PCHAS_0831300',
                                 'AMA1' = 'PCHAS_0931000')

cir_housekeeping_id <- c(cir_id, 
                         'PCHAS_0721000',
                         'PCHAS_0831300',
                         'PCHAS_0931000')

# Get order of pir genes by highest expressed genes in each sample.

combined_data_pir_ht <-  filter(chab_tpm_avg.df,
                             Geneid %in% cir_housekeeping_id) %>% 
  column_to_rownames(var = 'Geneid') %>% 
  filter_all(
    any_vars(. > 10)
  )

@@

```{r heatmap_df}
#Change the colnames to the times
colnames(combined_data_pir_ht) <- str_extract(colnames(combined_data_pir_ht),
                                              pattern = '.{3}_([[:digit:]]{2}h|sch.culture)')

#Order by time point
combined_data_pir_ht <- combined_data_pir_ht[,order(colnames(combined_data_pir_ht)),
                                             drop = FALSE]

#Add mean column for ordering
combined_data_pir_ht$mean_tpm <- apply(combined_data_pir_ht, 1, mean)

combined_data_pir_ht <- arrange(combined_data_pir_ht, desc(mean_tpm))

#Add a column for the subfam (ref for the reference genes)
combined_data_pir_ht$subfam <- ifelse(rownames(combined_data_pir_ht) %in% cir_id,
                                      cir_info$subfam_num[match(rownames(combined_data_pir_ht), 
                                                            cir_info$Geneid)],
                                      'ref')

#Get the ref gene actual names instead of the Geneid
rownames(combined_data_pir_ht) <- ifelse(
  rownames(combined_data_pir_ht) %in% housekeeping_genes[1,],
  colnames(housekeeping_genes)[ match(rownames(combined_data_pir_ht), housekeeping_genes[1,]) ],
  rownames(combined_data_pir_ht)
)

#Get the colour range
maxcol <- ceiling(
  max(
    log2(1+combined_data_pir_ht[!colnames(combined_data_pir_ht) %in% 'subfam'])
  ))

#Sub-family annotation
clade_info <- combined_data_pir_ht$subfam
annotation_col <- c(clade_col, 'ref' = 'black')

#Subfamily annotation

clade_ha <- rowAnnotation(
  'sub-family' = anno_simple(
    clade_info,
    col = annotation_col,
  ),
  annotation_name_gp = gpar(fontsize = 10)
)

#Subfamily name labels
num_subfam <- cir_info %>% 
  group_by(subfam) %>% 
  summarise(num = n_distinct(Gene.ID))

labels_subfam <- c(setNames(paste0(num_subfam$subfam, ' (n = ', num_subfam$num, ')'),
                            num_subfam$subfam),
                   'ref' = 'ref')
```

```{r draw_heatmap}
ht.list <- lapply(c('rMT', 'SBP'),
                  function(transmission){

                    ht.df <- combined_data_pir_ht[
                      ,
                      grepl(colnames(combined_data_pir_ht), 
                            pattern = paste0(transmission))
                    ]

                    colnames(ht.df) <- str_remove(colnames(ht.df),
                                                  pattern = paste0(transmission, 
                                                                   '_'))

                    ht <- Heatmap(
                      log2(
                        1+ht.df
                      ),
                      cluster_rows = FALSE,
                      cluster_columns = FALSE,
                      show_row_names = FALSE,
                      name = 'log(TPM+1)', 
                      col = colorRamp2(c(0,
                                         maxcol/4,
                                         maxcol/2,
                                         3*maxcol/4,
                                         maxcol),
                                       viridis(5)),
                      na_col = 'white', 
                      border = TRUE,
                      border_gp = gpar(lwd = 0.1),
                      column_names_gp = gpar(fontsize = 10),
                      row_dend_width = unit(3, "cm"),
                      width = unit(4.5, 'cm'),
                      height = unit(12,'cm'),
                      column_title = paste(transmission),
                      row_title_rot = 0,
                      row_title_gp = gpar(fontsize = 10),
                      row_split = forcats::fct_shift(
                        relevel(
                          as.factor(combined_data_pir_ht$subfam),
                          ref = 'ref')
                      ),
                      heatmap_legend_param = list(
                        labels_gp = gpar(fontsize = 10),
                        title_gp = gpar(fontsize = 10,
                                        fontface = 'bold'),
                        at = c(0,
                               maxcol/4,
                               maxcol/2,
                               3*maxcol/4,
                               maxcol)
                      )
                    )
                    return(ht)
                  })

ht.list <- Reduce('+', ht.list)

clade_lgd <- Legend(
  title = 'sub-family',
  at = c(labels_subfam),
  legend_gp = gpar(fill = annotation_col)
)

if(global_save_var){
  pdf(file = 'plots/subfam_TPMheatmap.pdf',
      height = 8,
      width = 6)
  draw(ht.list, 
       row_title_side = 'left',
       column_title_side = 'bottom',
       heatmap_legend_side = 'right',
       merge_legend = TRUE)
  dev.off()
}
```

```{r chapl_aapl_heatmap}

combined_data_pir_ht$locus <- cir_info$chapl_aapl_name_num[match(rownames(combined_data_pir_ht), 
                                    cir_info$Gene.ID)]

#Remove the pirs in the 'NA' locus (i.e. not in a chapl or aapl)
# don't remove actual NAs because these are the ref genes.
combined_data_pir_locus_ht <- filter(combined_data_pir_ht,
                                     !grepl(locus, pattern = "NA \\(n" ))

combined_data_pir_locus_ht$locus <- ifelse(is.na(combined_data_pir_locus_ht$locus),
                                     'ref',
                                     combined_data_pir_locus_ht$locus)

#Change locus to factor and change levels so that AAPLs are before ChAPLs and are together.
combined_data_pir_locus_ht$locus <- factor(combined_data_pir_locus_ht$locus,
                                           levels = ordering(combined_data_pir_locus_ht$locus,
                                                             order_vector = c('AAPL',
                                                                              'ChAPL',
                                                                              'ref')))

clade_info <- getElement(cir_info,
                         'subfam')[match(rownames(combined_data_pir_locus_ht), 
                                                 cir_info$Geneid)]
clade_info[is.na(clade_info)] <- 'ref'
names(clade_info) <- rownames(combined_data_pir_locus_ht)
clade_ha <- rowAnnotation(
  'subfamily' = anno_simple(
    clade_info,
    col = annotation_col
  )
)
#Draw the clade legend for the subfamily
clade_lgd <- Legend(
  title = 'sub-family',
  at = names(annotation_col),
  legend_gp = gpar(fill = annotation_col)
)

#Get the colour range
maxcol <- ceiling(
  max(
    log2(1+combined_data_pir_locus_ht[
      !colnames(combined_data_pir_locus_ht) %in% c('subfam', 
                                                   'locus')
    ])
  ))
```

```{r draw_locus_heatmap}

ht.list <- lapply(c('rMT', 'SBP'),
                  function(transmission){

                    #Get either the rMT or the SBP data
                    ht.df <- combined_data_pir_locus_ht[
                      ,
                      grepl(colnames(combined_data_pir_locus_ht), 
                            pattern = paste0(transmission))
                    ]

                    #Remove the transmission name from the column names
                    colnames(ht.df) <- str_remove(colnames(ht.df),
                                                  pattern = paste0(transmission, 
                                                                   '_'))

                    ht <- Heatmap(
                      log2(
                        1+ht.df
                      ),
                      cluster_rows = FALSE,
                      cluster_columns = FALSE,
                      show_row_names = FALSE,
                      name = 'log(TPM+1)', 
                      col = colorRamp2(c(0,
                                         maxcol/4,
                                         maxcol/2,
                                         3*maxcol/4,
                                         maxcol),
                                       viridis(5)),
                      na_col = 'white', 
                      border = TRUE,
                      border_gp = gpar(lwd = 0.1),
                      column_names_gp = gpar(fontsize = 10),
                      row_dend_width = unit(3, "cm"),
                      width = unit(4.5, 'cm'),
                      height = unit(12,'cm'),
                      column_title = paste(transmission),
                      row_title_rot = 0,
                      row_title_gp = gpar(fontsize = 10),
                      row_split = forcats::fct_shift(
                        relevel(
                          as.factor(combined_data_pir_locus_ht$locus),
                          ref = 'ref')
                      ),
                      heatmap_legend_param = list(
                        labels_gp = gpar(fontsize = 10),
                        title_gp = gpar(fontsize = 10,
                                        fontface = 'bold'),
                        at = c(0,
                               maxcol/4,
                               maxcol/2,
                               3*maxcol/4,
                               maxcol)
                      ),
                      #Only want the clade annotation on one heatmap not both.
                      left_annotation = if(transmission == 'rMT'){clade_ha}else{NULL}
                    )
                    return(ht)
                  })

ht.list <- Reduce('+', ht.list)

if(global_save_var){
  pdf(file = 'plots/ChAPL-AAPL_TPMheatmap.pdf',
      height = 9,
      width = 6)
  draw(ht.list, 
       row_title_side = 'left',
       column_title_side = 'bottom',
       heatmap_legend_side = 'right',
       merge_legend = TRUE,
       annotation_legend_list = list(clade_lgd))
  dev.off()
}

>>>>

library('readr')
library('dplyr')
library('ggplot2')
library("RColorBrewer")
library('data.table')
library('tibble')
library('stringr')
library('kableExtra')
library('viridis')
library('reshape2')
library('ComplexHeatmap')
library('circlize') # for circular plots and colour
library("factoextra")
library("MatrixGenerics") # for rowMins and rowMax
library('qusage')

@@

cir_info <- read_delim('../standard_files/Pchabaudi_cir_info.txt',
           delim = '\t')
cir_id <- cir_info$Geneid

cir_clade_info <- read_delim('../standard_files/Pchabaudi_pir_clades.txt',
            delim = '\t')
```

@@

# chab_tpm_avg.list <- list(
#   'RMT' = list(
#     tpm = chab_tpm_avg.df[,grepl(colnames(chab_tpm_avg.df), 
#                                 pattern = 'rMT|Geneid')],
#     experiment = 'RMT'
#   ),
#   'SBP' = list(
#     tpm = chab_tpm_avg.df[,grepl(colnames(chab_tpm_avg.df), 
#                                 pattern = 'SBP|Geneid')],
#     experiment = 'SBP'
#   )
# )

@@

```{r df}
#Filter low expressed genes (greater than 10 tpm in at least one sample), convert to a matrix, add 1 for the log transformation, then transform to log2
chab_tpm.df <-column_to_rownames(chab_tpm.df, var = 'Geneid')
chab_tpm.df <- chab_tpm.df[apply(chab_tpm.df, 1, function(x) {any(x >= 10)}),]
chab_tpm_log2plus1.mat <- chab_tpm.df %>% 
  as.matrix %>% 
  `+`(.+1) %>% 
  log(base = 2)
chab_tpm_log2plus1.mat[1:5,1:5]
```

```{r generate_labels}
#Use make.names to remove the '-' from the labels
colnames(chab_tpm_log2plus1.mat) <- make.names(colnames(chab_tpm_log2plus1.mat))
```

Removing genesets with < 3 members.

```{r generate_genesets}
pir_geneset <- list(pirs = cir_info$Geneid)

subfam_list <- lapply(unique(cir_info$subfam), 
                      function(clade_nam) filter(cir_info, 
                                                 subfam == clade_nam)$Geneid)
names(subfam_list) <- unique(cir_info$subfam)
subfam_list <- subfam_list[unlist(lapply(subfam_list, 
                                         function(x) length(x) > 3))]

ShortLong_list <- 
  lapply(unique(cir_info$SorL), 
         function(SorL_nam) filter(cir_info, 
                                   SorL == SorL_nam)$Geneid)
names(ShortLong_list) <- unique(cir_info$SorL)
ShortLong_list <- ShortLong_list[unlist(lapply(ShortLong_list, 
                                               function(x) length(x) > 3))]
order_path.names <- ordering(c('pirs', unique(cir_info$SorL), sort(unique(cir_info$subfam))), 
                             order_vector = c('pir','L','S'))

locus_list <- 
  lapply(unique(cir_info$chapl_aapl_name), 
         function(chapl_aapl_name_nam) filter(cir_info, 
                                   chapl_aapl_name == chapl_aapl_name_nam)$Geneid)
names(locus_list) <- unique(cir_info$chapl_aapl_name)
locus_list <- locus_list[unlist(lapply(locus_list, 
                                               function(x) length(x) > 3))]

@@

Which comparisons do I want to make?

```{r running_qusage}
ordered_stages <- ordering(unique(lifeCycle_labels),
         order_vector = times_stages_table$time)
#Remove sch.culture from the comparisons for now.
ordered_stages <- ordered_stages[!ordered_stages == 'sch.culture']
comparisons <- c(
  unlist(
    lapply(seq_along(ordered_stages),
           function(i){ 
             if(i < length(ordered_stages)){ 
               paste0(ordered_stages[i+1],'-',ordered_stages[i]) 
             } else {
               #When at the end, want it to cycle back round
               paste0(ordered_stages[1],'-',ordered_stages[i]) 
             } 
           }
    )
  )
)
```

Compare the gene lists between the same stages in different transmission

```{r qusage_stagesBetweenTransmission}

comparisons_stagesTrans <- paste0(paste('rMT', 
                                        times_stages_table$time, 
                                        sep = '_'),
                                  '-',
                                  paste('SBP', 
                                        times_stages_table$time, 
                                        sep = '_'))

comp_stagesTrans.mat <- chab_tpm_log2plus1.mat[
  ,!grepl(colnames(chab_tpm_log2plus1.mat), pattern = 'sch.culture')
]
# For the life cycle labels then extract the transmission + timepoint label
lifeCycle_labels <- str_extract(colnames(comp_stagesTrans.mat),
                                pattern = '^(rMT|SBP)_([[:digit:]]{2}h|sch.culture)')

all_comparisons <- lapply(comparisons_stagesTrans, 
                          function(comparison){
                            print(comparison)
                            list_comp <- lapply(
                              list(pir_geneset,
                                   ShortLong_list,
                                   subfam_list,
                                   locus_list), 
                              function(gene_list){
                                print(deparse(substitute(gene_list)))
                                qs.results.subfam <- qusage(comp_stagesTrans.mat,
                                                            lifeCycle_labels,
                                                            comparison,
                                                            gene_list,
                                                            n.points = 2^18)
                                list(results = qs.results.subfam, 
                                     qstable = qsTable(qs.results.subfam) %>% 
                                       mutate(contrast = qs.results.subfam$contrast))
                              })
                            names(list_comp) <- c('pir', 'shortlong', 'subfam', 'locus')
                            list_comp$combined <- bind_rows(lapply(list_comp, 
                                                                   getElement, 
                                                                   'qstable'))
                            return(list_comp)
                          })
names(all_comparisons) <- comparisons

all_comparisons.df <- bind_rows(lapply(all_comparisons, 
                                       function(x) bind_rows(lapply(x, 
                                                                    function(y) y$qstable)))) %>% 
  arrange(pathway.name)

stagesTrans_ht.mat <- filter(all_comparisons.df, FDR < 0.05) %>% 
  mutate(pathway.name = factor(pathway.name, 
                           levels = ordering(pathway.name, 
                                             order_vector = c('pirs', 'L', 'S', 'AAPL', 'ChAPL')))) %>% 
  arrange(pathway.name) %>% 
  melt(id.vars = c('contrast', 'pathway.name'),
       measure.vars  = c('log.fold.change')) %>% 
  dcast(contrast ~ pathway.name, 
        value.var = 'value') %>% 
  column_to_rownames('contrast') %>% 
  as.matrix

maxcol <- my_max(stagesTrans_ht.mat)
mincol <- my_max(stagesTrans_ht.mat*-1)*-1

ht <- Heatmap(stagesTrans_ht.mat,
        cluster_rows = FALSE,
        cluster_columns = FALSE,
        cluster_column_slices = FALSE,
        na_col = 'white',
        column_split = factor(str_extract(colnames(stagesTrans_ht.mat),
                                   pattern = 'pirs|^L|^S|AAPL|ChAPL'),
                              levels = unique(str_extract(colnames(stagesTrans_ht.mat),
                                   pattern = 'pirs|^L|^S|AAPL|ChAPL'))),
        column_order = colnames(stagesTrans_ht.mat),
        column_gap = unit(.5,"cm"),
        border = TRUE,
        col = colorRamp2(c(mincol,
                           0,
                           maxcol/4,
                           maxcol/2,
                           maxcol),
                         c('dodgerblue1', 'white', 'yellow', 'orange', 'red')),
        heatmap_legend_param = list(title = 'LogFoldChange'))

if(global_save_var){
  pdf(file = 'plots/stageTrans_logFoldheatmap.pdf',
      height = 8,
      width = 6)
  draw(ht)
  dev.off()
}
